Estimating the replicability of psychology experiments after an initial failure to replicate

### 2024-04-03 manuscript/manuscript.Rmd
typos and missing bib entries

---
title: "Estimating the replicability of psychology experiments after an initial failure to replicate"
author:
  - name: Veronica Boyce
    affiliation: Psych
    footnote:
      - corresp
  - name: Ben Prystawski
    affiliation: Psych
  - name: Adani Abutto
    affiliation: Psych
  - name: Emily Chen
    affiliation: Psych
  - name: Ziwen Chen
    affiliation: GSB
  - name: Howard Chiu
    affiliation: Edu
  - name: Irmak Ergin
    affiliation: Psych
  - name: Anmol Gupta
    affiliation: Psych
  - name: Chuqi Hu
    affiliation: Symsys
  - name: Bendix Kemmann
    affiliation: Philosophy
  - name: Nastasia Klevak
    affiliation: Psych
  - name: Verity Y. Q. Lua
    affiliation: Psych
  - name: Mateus M. Mazzaferro
    affiliation: Edu
  - name: Khaing Mon
    affiliation: Symsys
  - name: Dan Ogunbamowo
    affiliation: Psych
  - name: Alexander Pereira
    affiliation: Philosophy
  - name: Jordan Troutman
    affiliation: CS
  - name: Sarah Tung
    affiliation: Psych
  - name: Raphael Uricher
    affiliation: Psych
  - name: Michael C. Frank
    affiliation: Psych
address:
  - code: Psych
    address: Department of Psychology, Stanford University
  - code: GSB
    address: Graduate School of Business, Stanford University
  - code: Edu
    address: Graduate School of Education, Stanford University
  - code: Symsys
    address: Symbolic Systems Program, Stanford University
  - code: Philosophy
    address: Department of Philosophy, Stanford University
  - code: CS
    address: Department of Computer Science, Stanford University
footnote:
  - code: corresp
    text: "Corresponding author. Email: vboyce@stanford.edu"
bibliography: ["251rescue.bib"] # Replace with one or more of your own bibtex files. Better BibTeX for Zotero is your friend
csl: apa7.csl # Use any CSL style. See https://www.zotero.org/styles for a good list. Ignored if citation_package: natbib
link-citations: TRUE
output:
  bookdown::pdf_document2:
    toc: FALSE
    keep_tex: TRUE
    template: generic_article_template.tex
    #md_extensions: "-autolink_bare_uris"
    number_sections: TRUE
    citation_package: default # Can also be "natbib"
lang: en # Main document language in BCP47 format
geometry: "margin=25mm"
papersize: a4
#linestretch: 2 # for double spacing
endfloat: FALSE # Set to TRUE to turn on latex endfloat package to place figures and tables at end of document
# endfloatoption: # See endfloat documentation for more possibilities
numberlines: FALSE
authblk: TRUE # FALSE = author affiliations in footnotes; TRUE = author affiliations in a block below author names
footnotehyper: FALSE # TRUE will give you enhanced table footnote capabilities. Set to FALSE to be able to use French blocks. Needed due to what appears to be a latex bug.
urlcolor: blue
linkcolor: blue
citecolor: blue
graphics: TRUE # Needed to be able to include images
tables: TRUE # Needed to be able to include tables
# fancyhdr:
#   first:
#     #headleft: "REPORT-NO-XXXX"
#     headright: "Kaplan et al. (2021)"
#     headrulewidth: 0pt
#     #footleft: A left foot
#     footrulewidth: 0pt
#   subsequent:
#     #headleft: "NEXT-PAGE-HEADER-LEFT"
#     headright: "Kaplan et al. (2021)"
#     headrulewidth: 1pt
#     footrulewidth: 0pt
header-includes:
 - \usepackage{tikz}
 - \usetikzlibrary{positioning,chains}
 - \usepackage{setspace}\singlespacing
 - \renewcommand{\textfraction}{0.00}
 - \renewcommand{\topfraction}{1}
 - \renewcommand{\bottomfraction}{1}
 - \renewcommand{\floatpagefraction}{1}
 - \setcounter{topnumber}{3}
 - \setcounter{bottomnumber}{3}
 - \setcounter{totalnumber}{4}
editor_options: 
  markdown: 
    wrap: sentence
---

@@

For the goals of large-scale replication projects, it doesn't matter why any particular study failed to replicate, because the aim is to estimate the proportion of replication effects in a literature. In contrast, howver, when individual scientists attempt to replicate an effect with the goal of building on it, they do care about why an individual replication failed. In an uncertain literature, scientists might not want to commit resources to studying extensions of an effect without first checking that they can measure that effect themselves. For those individual scientists, it is much more important whether the replication failure is due to a fixable issue with the the replication or an intrinsic feature of the original result.

@@

Our analysis plan was pre-registered after students had selected projects, but before final data collection on the projects.
Each project was also individually pre-registered by the student conducting it.
The overall analysis is pre-registered at https://osf.io/5qz7v, individual pre-registrations are linked from TODO LINK.
We note one deviation from our pre-registration here: we pre-registered visual comparisons between original, first replication, and rescue projects using prediction intervals.
Prediction intervals depend on both the original effect size and variance and the variance of the comparison (replication or rescue) study [@patil2016].
Thus we cannot show a single prediction interval for the original study, but would have to show a prediction interval between each pair of studies. We thought this analysis would not offer additional clarity and hence we did not perform it.

@@

Students were required to think of reasons the first replication might not have worked, and address them if they could.
A list of possible reasons and solutions TODO LINK was given to students.
In general, we encouraged students to add manipulation checks as appropriate, and to adapt their materials to online studies, which may benefit from attention checks and/or shorter duration.
For instance, the rescue of @paxton2012 switched from using the cognitive reflection task (CRT) which many online participants are over-familiar with for the newer CRT-2 which is less overused but attempts to measure the same construct [@thomson2016].
The rescue of @jara-ettinger2022 discovered that the replication had accidentally used an illustrated version of the stimuli rather than the photographic stimuli used in the original and used the photographic stimuli instead.
@tarampi2016 had participants do a timed navigation task where they wrote L or R to indicate which way to turn at each intersection on a paper map. The replication ported this experiment online and had participants click a drop down menu to select left or right for each turn. The rescue removed the unwieldy drop down menus and opted to have participants press keys on the keyboard to indicate the direction of a turn, which seemed like a more natural interface.
Once experimental designs and analytic plans were approved by class TAs (VB and BP), students pre-registered and ran their samples.

@@

@ngo2019 was a second successful rescue. The original study found a large effect; hence the first replication, powering for 80% power on the reported effect, recruited a small sample of 12 people. This small study failed to find the original effect.
The rescue, powered using 2.5x the original sample [as recommended by @simonsohn2015], recovered a clear effect (albeit a much smaller one).
There are reasons to think that some effect sizes in the literature may be inflated [@openscienceconsortium, @camerer2018], and it is also possible that changes to experiments or switches to online could result in noisier samples and thus smaller effect sizes.
Therefore, replications with smaller samples than the original (even if powered to the original effect size), may not be very diagnostic, and could potentially benefit from a re-replication.

>>>>

---
title: "Estimating the replicability of psychology experiments after an initial failure to replicate"
author:
  - name: Veronica Boyce
    affiliation: Psych
    footnote:
      - corresp
  - name: Ben Prystawski
    affiliation: Psych
  - name: Adani Abutto
    affiliation: Psych
  - name: Emily Chen
    affiliation: Psych
  - name: Ziwen Chen
    affiliation: GSB
  - name: Howard Chiu
    affiliation: Edu
  - name: Irmak Ergin
    affiliation: Psych
  - name: Anmol Gupta
    affiliation: Psych
  - name: Chuqi Hu
    affiliation: Symsys
  - name: Bendix Kemmann
    affiliation: Philosophy
  - name: Nastasia Klevak
    affiliation: Psych
  - name: Verity Y. Q. Lua
    affiliation: Psych
  - name: Mateus M. Mazzaferro
    affiliation: Edu
  - name: Khaing Mon
    affiliation: Symsys
  - name: Dan Ogunbamowo
    affiliation: Psych
  - name: Alexander Pereira
    affiliation: Philosophy
  - name: Jordan Troutman
    affiliation: CS
  - name: Sarah Tung
    affiliation: Psych
  - name: Raphael Uricher
    affiliation: Psych
  - name: Michael C. Frank
    affiliation: Psych
address:
  - code: Psych
    address: Department of Psychology, Stanford University
  - code: GSB
    address: Graduate School of Business, Stanford University
  - code: Edu
    address: Graduate School of Education, Stanford University
  - code: Symsys
    address: Symbolic Systems Program, Stanford University
  - code: Philosophy
    address: Department of Philosophy, Stanford University
  - code: CS
    address: Department of Computer Science, Stanford University
footnote:
  - code: corresp
    text: "Corresponding author. Email: vboyce@stanford.edu"
bibliography: ["251rescue.bib"] # Replace with one or more of your own bibtex files. Better BibTeX for Zotero is your friend
csl: apa7.csl # Use any CSL
nocite: |
  @krauss2003, @ngo2019, @todd2016, @jara-ettinger2022, @porter2016, @birch2007, @child2018, @chou2016, @craig2014, @gong2019, @haimovitz2016, @hopkins2016, @paxton2012, @payne2008, @schechtman2010, @tarampi2016, @yeshurun2003
link-citations: TRUE
output:
  bookdown::pdf_document2:
    toc: FALSE
    keep_tex: TRUE
    template: generic_article_template.tex
    #md_extensions: "-autolink_bare_uris"
    number_sections: TRUE
    citation_package: default # Can also be "natbib"
lang: en # Main document language in BCP47 format
geometry: "margin=25mm"
papersize: a4
#linestretch: 2 # for double spacing
endfloat: FALSE # Set to TRUE to turn on latex endfloat package to place figures and tables at end of document
# endfloatoption: # See endfloat documentation for more possibilities
numberlines: FALSE
authblk: TRUE # FALSE = author affiliations in footnotes; TRUE = author affiliations in a block below author names
footnotehyper: FALSE # TRUE will give you enhanced table footnote capabilities. Set to FALSE to be able to use French blocks. Needed due to what appears to be a latex bug.
urlcolor: blue
linkcolor: blue
citecolor: blue
graphics: TRUE # Needed to be able to include images
tables: TRUE # Needed to be able to include tables
# fancyhdr:
#   first:
#     #headleft: "REPORT-NO-XXXX"
#     headright: "Kaplan et al. (2021)"
#     headrulewidth: 0pt
#     #footleft: A left foot
#     footrulewidth: 0pt
#   subsequent:
#     #headleft: "NEXT-PAGE-HEADER-LEFT"
#     headright: "Kaplan et al. (2021)"
#     headrulewidth: 1pt
#     footrulewidth: 0pt
header-includes:
 - \usepackage{tikz}
 - \usetikzlibrary{positioning,chains}
 - \usepackage{setspace}\singlespacing
 - \renewcommand{\textfraction}{0.00}
 - \renewcommand{\topfraction}{1}
 - \renewcommand{\bottomfraction}{1}
 - \renewcommand{\floatpagefraction}{1}
 - \setcounter{topnumber}{3}
 - \setcounter{bottomnumber}{3}
 - \setcounter{totalnumber}{4}
editor_options: 
  markdown: 
    wrap: sentence
---

@@

For the goals of large-scale replication projects, it doesn't matter why any particular study failed to replicate, because the aim is to estimate the proportion of replication effects in a literature. In contrast, however, when individual scientists attempt to replicate an effect with the goal of building on it, they do care about why an individual replication failed. In an uncertain literature, scientists might not want to commit resources to studying extensions of an effect without first checking that they can measure that effect themselves. For those individual scientists, it is much more important whether the replication failure is due to a fixable issue with the the replication or an intrinsic feature of the original result.

@@

Our analysis plan was pre-registered after students had selected projects, but before final data collection on the projects.
Each project was also individually pre-registered by the student conducting it.
The overall analysis is pre-registered at https://osf.io/5qz7v, individual pre-registrations are linked from https://osf.io/vkbfw.
We note one deviation from our pre-registration here: we pre-registered visual comparisons between original, first replication, and rescue projects using prediction intervals.
Prediction intervals depend on both the original effect size and variance and the variance of the comparison (replication or rescue) study [@patil2016].
Thus we cannot show a single prediction interval for the original study, but would have to show a prediction interval between each pair of studies. We thought this analysis would not offer additional clarity and hence we did not perform it.

@@

Students were required to think of reasons the first replication might not have worked, and address them if they could.
A list of possible reasons and solutions was given to students (https://osf.io/dfwk2).
In general, we encouraged students to add manipulation checks as appropriate, and to adapt their materials to online studies, which may benefit from attention checks and/or shorter duration.
For instance, the rescue of @paxton2012 switched from using the cognitive reflection task (CRT) which many online participants are over-familiar with for the newer CRT-2 which is less overused but attempts to measure the same construct [@thomson2016].
The rescue of @jara-ettinger2022 discovered that the replication had accidentally used an illustrated version of the stimuli rather than the photographic stimuli used in the original and used the photographic stimuli instead.
@tarampi2016 had participants do a timed navigation task where they wrote L or R to indicate which way to turn at each intersection on a paper map. The replication ported this experiment online and had participants click a drop down menu to select left or right for each turn. The rescue removed the unwieldy drop down menus and opted to have participants press keys on the keyboard to indicate the direction of a turn, which seemed like a more natural interface.
Once experimental designs and analytic plans were approved by class TAs (VB and BP), students pre-registered and ran their samples.

@@

@ngo2019 was a second successful rescue. The original study found a large effect; hence the first replication, powering for 80% power on the reported effect, recruited a small sample of 12 people. This small study failed to find the original effect.
The rescue, powered using 2.5x the original sample [as recommended by @simonsohn2015], recovered a clear effect (albeit a much smaller one).
There are reasons to think that some effect sizes in the literature may be inflated [@openscienceconsortium; @camerer2018], and it is also possible that changes to experiments or switches to online could result in noisier samples and thus smaller effect sizes.
Therefore, replications with smaller samples than the original (even if powered to the original effect size), may not be very diagnostic, and could potentially benefit from a re-replication.
